Extrinsic Type Inference (DRAFT 2025-06-08)

### 2025-07-09 extrinsic_type_inference_paper/paper.tex
udpate related work

In order to type the program \code{fooPair}, a type system would either need to determine that
\code{f} is parametric, or that the types of \code{123} and \code{true} are compatible.
Treating the parameter \code{f} is allowed in systems that have impredicative parametric polymorphism \cite{}, 
which was introduced by System F \cite{}, 
introduced into ML-style languages with boxed polymorphism \cite{} 
and developed further with ML$^F$ \cite{} and various other systems \cite{}. 
Unfortunately, it's not clear what parametric type should be inferred for \code{f}.
For instance, \code{forall 'a . 'a -> 'a}, \code{forall 'a . 'a -> 'a list}, \code{forall 'a . 'a -> 'a option}, 
or an infinite number of
other parametric types could be used, but none is weaker than all the rest.
Thus, these systems with impredicative parametric polymorphism rely on annotations of parameters to
determine second order parametric types, which constitutes intrinsic typing of
of functions. 
Since there is no parameter type that is weaker than all the rest,
merely relying on intrinsic second order parametric types 
significantly restricts the reusability of expressions.
Additionally, many of these techniques require explicit instantiation to use parametrically typed
expressions, which can be seen as an intrinsic restriction that 
delays pruning the control-flow until a particular usage.

To reduce the annotation burden and increase expressivity, researchers
developed ways to infer type instantiations and compare type with varying
amounts of possible instantiations, by combining impredicative types
with subtyping and type inference techniques. Bidirectional type inference \cite{}  
is one such technique where bounds on type variables can be inferred in 
both directions of the control flow.
It propagates inferred and annotated knowledge directly from
the structures of the program and relies on subtyping merely to check compatibility between
types. Additionally, whether a non-parametric type is an instantiation
of a parametric type is treated as distinct from subtyping.

To allow arbitrary mixing of constructions in pattern matching, the extrinsic type of constructions
must be abolished so that there is no intrinsic weakest type other than a top type. 
Typescript \cite{} does not have an intrinsic weakest type for constructors,
and it can infer output types of functions that return different structural forms in different branches.
However, it is unable to infer a weak type for the parameters, and thus requires annotation of parameters.  
Typescript and other systems for scripting languages, rely on flow-sensitive, or occurrence typing \cite{},
which can refine the types of variables in branches based on the condition guarding each branch.
For Typescript, this flow information is only used in one direction, and it is not leveraged to infer
the weakest type of parameters.

\TODO{mention algebraic subtyping, Cretin, and flow analysis}

\TODO{note how solving subtyping constraints subsume the notion of bidirection typig}

In order to avoid this loss of expressivity and the annotation burden of intrinsic typing, 
SuperF \cite{} introduced 
a technique that relies on extrinsic subtyping polymorphism, 
which bypasses the need for intrinsic parametric polymorphism.
In the previous example, SuperF can infer the principal type for \code{fooPair} by inferring the 
weakest type of its parameter \code{f}, which it achieved by leveraging subtyping to find 
a type that is compatible with both \code{123} and \code{true}.
The SuperF type system infers subtyping constraints that represent the data flow represented by
function application and pattern matching. It relies 
intersection and union type combinators to represent the strongest or weakest interpretations
of type variables that meet certain subtyping constraints.   
With SuperF's type inference mechanism, it is possible to infer the principal type for the 
previous problematic example \code{fooPair: (forall 'a 'b . (int -> 'a) /\textbackslash\ (bool -> 'b) -> 'a * 'b)}.

@@

@@

\TODO{discuss implicit and flow based approaches}

\TODO{enumerate the various work with various type forms}

\TODO{note how our structural requirements on recursive types prevent typing generators}

\TODO{note the distinction between our least fixed point types and the more general 
recursive types}

>>>>

In order to type the program \code{fooPair}, a type system would either need to determine that
\code{f} is parametric, or that the types of \code{123} and \code{true} are compatible.
Treating the parameter \code{f} is allowed in systems that have impredicative parametric polymorphism \cite{}, 
which was introduced by System F \cite{}.

To reduce the annotation burden and increase expressivity, researchers
developed ways to infer type instantiations and compare types with varying
multiple possible instantiations, by combining impredicative types
with type inference techniques and ways to compare approximations of varying precision .

One of possibility is bidirectional type inference \cite{}  
is one such technique where bounds on type variables can be inferred in 
both directions of the control flow.
It propagates inferred and annotated knowledge directly from
the structures of the program and relies on subtyping merely to check compatibility between
types. Additionally, whether a non-parametric type is an instantiation
of a parametric type is treated as distinct from subtyping.

In order to maintain HM's implicit first-order types in the context
of higher-order types, other techniques were explored.  
These techniques include boxed polymorphism \cite{}, ML$^F$ \cite{}, and various other systems \cite{}. 
Unfortunately, it's not clear what parametric type should be inferred for \code{f}.
For instance, \code{forall 'a . 'a -> 'a}, \code{forall 'a . 'a -> 'a list}, \code{forall 'a . 'a -> 'a option}, 
or an infinite number of
other parametric types could be used, but none is weaker than all the rest.
Thus, these systems with impredicative parametric polymorphism rely on annotations of parameters to
determine second order parametric types, which constitutes intrinsic typing of
of functions. 
Since there is no parameter type that is weaker than all the rest,
merely relying on intrinsic second order parametric types 
significantly restricts the reusability of expressions.
Additionally, many of these techniques require explicit instantiation to use parametrically typed
expressions, which can be seen as an intrinsic restriction that 
delays pruning the control-flow until a particular usage.

All the type inference techniques depend reasoning over the observed control flow  
of programs, but the reasoning to particular situations and objects and use different
relations for solving vs checking.  The subtyping relation was generalized in the theory of implicit coercions (Cretin) \cite{},
such that even instantiation is treated as subtyping (or implicit coercion).  
Based on these ideas, SuperF \cite{} introduced type inference techniques for generating actually solving
these generalized subtyping constraints

SuperF \cite{} introduced 
a technique that relies on extrinsic subtyping polymorphism, 
which bypasses the need for intrinsic parametric polymorphism.
In the previous example, SuperF can infer the principal type for \code{fooPair} by inferring the 
weakest type of its parameter \code{f}, which it achieved by leveraging subtyping to find 
a type that is compatible with both \code{123} and \code{true}.
The SuperF type system infers subtyping constraints that represent the data flow represented by
function application and pattern matching. It relies 
intersection and union type combinators to represent the strongest or weakest interpretations
of type variables that meet certain subtyping constraints.   
With SuperF's type inference mechanism, it is possible to infer the principal type for the 
previous problematic example \code{fooPair: (forall 'a 'b . (int -> 'a) /\textbackslash\ (bool -> 'b) -> 'a * 'b)}.

@@

\TODO{recursive types only matter in light of structures}

To allow arbitrary mixing of constructions in pattern matching, the extrinsic type of constructions
must be abolished so that there is no intrinsic weakest type other than a top type.

In MLSub and algebraic subtyping \cite{}, forgoes the datatype mechanism.
Instead, it uses type inference to generates subtyping constraints and solve them, akin to unification in HM.

\TODO{understand the capabilities and limitations of MLSub better}

Popular scripting languages have also been augmented with some type inference capabilities.
Typescript \cite{} does not have an intrinsic weakest type for constructions,
and it can infer output types of functions that return different structural forms in different branches.
However, it is unable to infer a weak type for the parameters, and thus requires annotation of parameters.  
Typescript and other systems for scripting languages, rely on flow-sensitive, or occurrence typing \cite{},
which can refine the types of variables in branches based on the condition guarding each branch.
For Typescript, this flow information is only used in one direction, and it is not leveraged to infer
the weakest type of parameters.

@@

\TODO{note work on recursive types and regular trees}
